Significant wind disturbances induced by giant dunes.

### 2021-12-02 Paper/main.tex
Re-ecriture jusqu'à section 3.1

Indeed, eolian sand dunes typically emerge from the interaction by the turbulent flow and the topography \citep{Charru2013}. Later on, when the dune reaches an intermediate size, it may induce large deflections that can significantly impact sediment pathways of coastal systems \citep{Hesp2015}. Likewise, the collective behavior of dune populations may also be affected by long-range interactions due to flow disturbances induced by each individual \citep{Smith2017, Bacik2020}. As the dunes increase in size by collisions and coarsening, they sometimes reach a giant size, comparable to the boundary layer depth, thus inducing interactions not only with the turbulent flow of the ABL, but also with the free atmosphere~\citep{andreotti2009}. However, the wind disturbances induced by these giant dunes have never been quantified.

@@

In this study, we focus on four places across and nearby the Namib desert, highlighting different environments (see Fig.~\ref{Fig1}).
%
The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, located on the coast at the outlet of the Huab river is an arid environment exhibiting $60$-m scale barchan dunes. While these two stations are in environments with no mid-scale topography, this is not the case for the Deep Sea and South Namib stations. Both are located in the interdune between giant linear dunes with kilometric wavelengths and more than $40$-m high. In this section, we describe and compare the wind regimes resulting from the available datasets in each station.

@@

  Finally, the dune properties are computed using an autocorrelation method from the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Namib and Deep Sea stations, we obtain respectively orientations of $85^\circ$ and $125^\circ$, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes of $45~\textup{m}$ and $20~\textup{m}$ (see SI section 2 for additional details).

@@

  The regional wind patterns in the Namib are strongly controlled by the see breeze, resulting in strong northward components present in all regional wind roses of figure~\ref{Fig1}, also deviated by the large scale topography \citep{lancaster1985}. These daily winds are the strongest during the second-half of the year (Septembre-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming on the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast, resulting in three bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}(c, d, e)) and one unidirectional wind regime on the coast at the outlet of the Huab River.

  In the case of the Adamax and Huab stations, these regional wind roses qualitatively matches their respective in situ local measurements. However, for the Deep Sea and South Namib stations, the local wind roses exhibit additional components aligned with the giant dune orientation visible on the satellite images.

  Indeed, the analysis of the wind speed and direction time series shows that the matching between the local and regional winds is verified for the stations with flat environment at all times (Fig.~\ref{Fig2}(a--b)), as well as for the stations within the giant dune field, but only during Septembre-January (Fig.~\ref{Fig2}(c--d). This agreement is statistically verified, as shown by the Fig.~\ref{Fig5_supp}.

@@

  In winter, the local and regional winds match only during the morning for the stations within the giant dune field, when the southerly/southwesterly sea breeze dominates (see Fig.~\ref{Fig2}(e--f), Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, the two datasets differs. More specifically, this mismatch mainly occurs during at low wind velocities. In this case, the angular wind distribution of the local measurements exhibits two additional modes separated of $\simeq 180^\circ$, each corresponding to the giant dune alignement (purple frame in Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). This deviation at low wind velocities is statistically verified, and is also associated with a global attenuation of the wind strength (Fig.~\ref{Fig6_supp} and Fig.~\ref{Fig7_supp}).

@@

  Previous studies have linked meteorological flow around and over obstacles to the vertical structure of the atmosphere \citep{Stull1988}. At the dune scale, its lower part, the turbulent atmospheric boundary layer (ABL), is directly influenced by the planetary surface \citep{Stull1988}. When the latter is warmer than the ambient air (typically during the day), a strong convection takes place, resulting in a logarithmic wind profile and a vertically constant potential temperature $\theta$ (which corresponds to the air density corrected from the vertical pressure variation). During the night, the surface cools faster than the ambient air, which can result in a stable layer close to the surface (typically the first $200$ meters). Above, the free atmosphere (FA) is a stably stratified zone in which turbulence is negligible, and where the flow is usually considered as incompressible and inviscid. In the middle, a transitional layer, also known as entrainment zone, is characterized by a sharp increase of the potential temperature, which traps turbulence resulting fro the surface friction below it. As it is much shallow than the other two layers, we consider it as capping singularity inducing a discontinuity in the potential temperature profile.

>>>>

Indeed, eolian sand dunes typically emerge from the feedback of the topography on the turbulent flow, which speeds up close to the dune crest \citep{Rubin1987, Charru2013, Courrech2014}. Later on, when the dune reaches an intermediate size, it may also induce significant wind deflections. This can impact the sediment pathways of coastal systems \citep{Hesp2015}, or affect the collective behavior of dune populations with long-range interactions due to flow disturbances induced by each individual \citep{Smith2017, Bacik2020}. As the dunes increase in size by collisions and coarsening, they sometimes reach a giant size, comparable to the boundary layer depth, thus inducing interactions not only with the turbulent flow of the ABL, but also with the free atmosphere~\citep{andreotti2009}. However, the wind disturbances induced by these giant dunes have never been quantified.

@@

In this study, we focus on four places across and nearby the Namib desert, highlighting different environments (see Fig.~\ref{Fig1}).
%
The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, located on the coast at the outlet of the Huab river is an arid environment exhibiting $60$-m scale barchan dunes. While these two stations are in environments with no mid-scale topography, this is not the case for the Deep Sea and South Namib stations. Both are located in the interdune between giant linear dunes with kilometric wavelengths and superimposed patterns. In this section, we describe and compare the wind regimes resulting from the available datasets in each station.

@@

  Finally, the dune properties are computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Namib and Deep Sea stations, we obtain respectively orientations of $85^\circ$ and $125^\circ$, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes of $45~\textup{m}$ and $20~\textup{m}$ (see Fig.~\ref{Fig4_supp} for more details).

@@

  The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daily winds are dominant during the second-half of the year (Septembre-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming on the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast, resulting in bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b, \ref{Fig1}d and \ref{Fig1}e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c).

  In the case of the Adamax and Huab stations, the regional wind roses qualitatively match those corresponding to the local in situ measurements. However, for the Deep Sea and South Namib stations, the local wind roses exhibit additional components aligned with the giant dune orientation visible on the satellite images (Fig.~\ref{Fig1}c--d).
  %
  Indeed, the analysis of the wind speed and direction time series shows that the agreement between the local and regional datasets is always verified when no mid-scale topography are present (Fig.~\ref{Fig2}a--b) and Fig.~\ref{Fig5_supp}). In contrast, for the stations within the giant dune field, we observe that this agreement is limited to the Septembre--January time periods (Fig.~\ref{Fig2}c--d).

@@

  When giant dunes are present, in the February--August period, the local and regional winds match only during the morning, i.e when the southerly/southwesterly sea breeze dominates (see Fig.~\ref{Fig2}(e--f), Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, the two datasets differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes separated of $\simeq 180^\circ$, each corresponding to the giant dune alignement (purple frame in Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}, as well as Fig.~\ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this process occurs for low wind velocities, typically for $u_{*} < 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $0.25~\textrm{m}~\textrm{s}^{-1}$, this wind reorientation does not occur. Finally, for intermediate shear velocities, both reorientation along the dune crest and no reorientation are observed (Fig.~\ref{Fig7_supp}).

@@

  In the case of linear ridges, dune-induced flow disturbances have mainly been related to the incident wind direction~\citep{Walker2009, Hesp2015}. In our case, it is unlikely to be the dominant parameter, as the most deflected wind for both stations is the most perpendicular, where it should be winds with a $30^{\circ}--70^{\circ}$ incident direction~\citep{Hesp2015}.
  An important observation is the difference in behavior between low and high wind velocities, which suggests a change in the hydrodynamical regime.

  Previous studies have linked meteorological flow around and over topographical obstacles to the vertical structure of the atmosphere \citep{Stull1988}. More particularly, dunes evolves in its lower part, the turbulent atmospheric boundary layer (ABL), typically characterized by a logarithmic wind profile and a vertically constant potential temperature $\theta$. Above, the free atmosphere (FA) is a stably stratified zone in which turbulence is negligible, and where the flow is usually considered as incompressible and inviscid. In the middle, a transitional layer, also known as entrainment zone, is characterized by a sharp increase of the potential temperature, which traps turbulence resulting from the surface friction below it.

  In the following, we sum-up the dominant numbers leading to different hydrodynamical interactions with the obstacle, and interpret the data with respect to the corresponding physical mechanisms.
